---
title: "Arsenic report"
author: "Randy L Coryell"
date: "June 17, 2016"
output: pdf_document
toc: true
toc_depth: 4
---


## Overview of Analysis

This report outlines a *regression* analysis of arsenic concentration as a function of well depth. it woll only include some basic assumption checks of normality and equal variance of the error terms.

**Note**: Arsenic is measured in ppb; well depth is measured in feet.


***

As we proceed in our analysis we shall endeavor to remember the following:

> The most that can be expected from any model is that it can supply a useful approximation to reality: All models are wrong; some models are useful.

>> *George Box*



***



## Read the Data

```{r read-data}
file.name <- "arsenic-03.data"
arsenic.data <- read.table(file=file.name, sep="\t", skip=5,
                       header=TRUE, na.strings=c("."))
dim(arsenic.data)
head(arsenic.data, 4)
```

The data has `r nrow(arsenic.data) -
sum(complete.cases(arsenic.data))` observations with 
missing values out of a total of 
`r nrow(arsenic.data)` observations.

## Analysis

### Visual Inspection of the Relationship

We wish to examine the data visually.

```{r fig.height=2.5, fig.width=4}
plot(arsenic ~ depth, data=arsenic.data, cex=0.3, pch="+", col="red")
```


```{r fig.height=4}
plot(arsenic ~ depth, data=arsenic.data, cex=.5, pch="o", col="red")
```

Perhaps a cubic relationship would fit the trend as it first curves upwards and then curves downwards.  It also appears that the variability increases in the middle.

We will try a regression fit of arsenic using a cubic polynomial in *depth*. We first create quadratic and cubic *depth* variables.

```{r}
arsenic.data <- within(arsenic.data, {
depth2 <- depth^2
depth3 <- depth^3
})
head(arsenic.data, 3)
reg01 <- lm(arsenic ~ depth + depth2 + depth3,
            data=arsenic.data)
summary(reg01)$coef
```


#### Table of Regression Coefficients

| Variable | Coefficient | Significant|
|:------------|------:|:-------:|
| Intercept     | 24.985 | No|
|Depth          | -6.217 | Yes|
|Depth2         | 0.408  | Yes|
|Depth3         | 0.005  | Yes|


We add the fitted curve to the data plot to see how 
well it appears to fit.

```{r fig.height=4.5}
plot(arsenic ~ depth, data=arsenic.data)
x.range <- range(arsenic.data$depth)
x <- seq(x.range[1], x.range[2], length.out=100)
curve.data <- data.frame(depth=x, depth2=x^2, depth3=x^3)
curve.data$arsenic <- predict(reg01, curve.data)
lines(x=curve.data$depth, y=curve.data$arsenic,
      col="blue", lwd=3)
```

The fit looks fairly decent.


### Assumptions

We now check assumptions about:

- error variances
    - should be equal
- error distribution
    - should be normal
    
We will do this with:

1. plotting $e_i$ versus $\hat{y}_i$
2. normal quantile plot


#### Equal Variance of Errors

We visually investigate the assumption of equal variances.


```{r fig.height=4.5}
resid <- residuals(reg01)
y.hat <- predict(reg01, arsenic.data)
plot(resid ~ y.hat)
abline(h=0)
```

```{r fig.height=4.5}
plot(resid ~ arsenic.data$depth)
abline(h=0)
```

Clearly the assumption of equal variances of the error terms is not met.

#### Normality of Errors

We visually investigate the assumption of normality.

```{r fig.height=4.5}
qqnorm(resid)
qqline(resid)
```


Clearly the normality assumption on the error terms is also not met.



### Transformation of the Response

To address the equal variance and normality of error term assumption violations, we attempt to find a transformation for the response variable using the BoxCox transformation method.

```{r fig.height=4.5}
library("MASS")
boxcox(reg01)
```

From the plot we see that the BoxCox technique points to a log-transformation (since $\lambda = 0$).



### Fit of Transformed Response

We try fitting the data with the log-transformed response, and then perform assumption checks again.

```{r fig.height=4.5}
arsenic.data$log.arsenic <- log(arsenic.data$arsenic)
reg02 <- lm(log.arsenic ~ depth + depth2 + depth3, data=arsenic.data)
resid <- residuals(reg02)
y.hat <- predict(reg02, arsenic.data)
plot(resid ~ y.hat)
abline(h=0)
plot(resid ~ arsenic.data$depth)
abline(h=0)
```

There does not appear to be any equal variance assumption violation with the log-transformed response.

>

We examine a plot of the log-transformed response versus depth.

```{r fig.height=4.5}
plot(log.arsenic ~ depth, data=arsenic.data)
```

The data appear much more consistent and looks as though a quadratic fit may be adequate.


### Examination of the Fit with Transformed Response

We examine the regression fit using the transformed response.

```{r}
summary(reg02)
```

>

Indeed, the cubic term is not significant, so we refit the model without the cubic term.

```{r}
reg03 <- lm(log.arsenic ~ depth + depth2, data=arsenic.data)
summary(reg03)
```

And the quadratic fit has all the terms highly significant (p-value < 0.0001).

>

We now plot the fit with the data.

```{r fig.height=4.5}
plot(log.arsenic ~ depth, data=arsenic.data)
x <- seq(min(arsenic.data$depth), max(arsenic.data$depth),
             length.out=100)
curve.data$log.arsenic <-predict(reg03, curve.data)
lines(x=curve.data$depth, y=curve.data$log.arsenic,
      col="blue", lwd=3)
```
